Left-linear rules in adhesive categories

### 2024-02-12 icalp/fusione.tex
abstract and intro

\author{Paolo Baldan} 
{Department of Mathematics, University of Padova, Italy}
{baldan@math.unipd.it}{}{}
\author{Davide Castelnovo}
{Department of Mathematics, University of Padova, Italy}
{davide.castelnovo@math.unipd.it}{}{}
\author{Andrea Corradini}
{Department of Computer Science, University of Pisa, Italy}
{andrea.corradini@unipi.it}{}{}

\author{Fabio Gadducci}
{Department of Computer Science, University of Pisa, Italy}
{fabio.gadducci@unipi.it}{}{}

@@

\begin{abstract}
  When two sequences of steps executed by a computational device can
  be considered equivalent?  This is a relevant question if one
  considers concurrent and distributed systems, since the independence
  of two steps means that they could be executed in any order, and
  potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  In fact, the standard characterisation of independence does not hold anymore,
  and this requires a substantial revision of the 
  theory for the linear case, since a number of basic properties,
  essential in the development of a theory of concurrency, were
  derived from such a description.
  %
  The paper introduce a novel characterisation of independence, for which we prove a Church-Rosser property,
  properly extending the class of formalisms 
  that can be modelled in the adhesive framework. 
\end{abstract}

@@

\item \emph{Tame rewriting systems}\\ An
  independence pairs induced by a filler is called \emph{good} A
  rewriting system where all indepedence pairs are good is called
  \emph{tame}

  NOTE: the categories in which
  condition $B$ from~\cite{baldan2011adhesivity} identifies a large
  class of (quasi)adhesive categories where left-linear systems are all
  tame. Adapted here to M-adhesivity.

\item \emph{Very Tame rewriting
    systems}\\ Tame and existing indepence pairs are unique (hence
  switching is unique) (see later for examples)
\end{itemize}

@@

This part is dealt with for very tame rewriting systems and depends on this assumption. Note that
\begin{itemize}
\item Linear rewriting systems on adhesive categories are very tame
\item Left-linear rewritint systems on graph-like categories with generalisation of (i) no isolated nodes in $L$ and (ii) right leg injective on nodes is very tame
\item Graphs with equivalences (Egg) with left-leg regular mono, right leg mono is very tame
\end{itemize}

@@

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required only to consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagram mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

% =========DA QUI SUPERSEDED\\
% Rewriting systems over graphs are used in uncountably many setting for
% representing the dynamics of software systems: the graph serves to represent
% the entities and their relations within states , while rewriting steps,
% modifying the graph, model computational steps that results in
% changes to the system state.

% Rewriting over many kind of graphs is elegantly captured by the
% so-called \emph{double pushout} (DPO) approach to rewriting. A
% noteworthy advantage is that DPO rewriting can be formulated over
% adhesive categories and their
% variants~\cite{lack2005adhesive,ehrig2006weak}. Operating within the
% framework of adhesive categories allows one to devise the foundations
% of a rewriting theory which can be subsequently instantiated to the
% context of interest, thus avoiding the need of repeatedly proving
% similar ad-hoc results in for each specific setting.

% Rewriting over adhesive categories has been widely studied in the last
% two decades and most of the research focused on the so-called linear
% rewriting rules. Roughly speaking, linear rewriting rules can specify
% that, in a rewriting step, some entities in the current state are
% removed, some other are needed for the rewriting step to occur but
% remain unaltered and some new entities can be generated.

% The computation described by a rewriting system can be naturally
% interpreted as a distributed computation and a notion of independence
% between rewriting steps naturally emerges. When different rules are
% applied at ``distinct'' parts of the state, possibly overlapping on
% preserved items, the rewriting steps can be naturally deemed as
% independent and their application can occur in any order and possibly
% concurrently. This phenomenon is formally captured by the so-called
% Church-Rosser theorem of DPO rewriting~\cite{CMREHL:AAGT}.
% %
% Having a notion of independence between computation steps sits at the
% core of the possibility of viewing an execution trace as a concurrent
% computation. This is obtained by considering traces up to an
% equivalence that identifies traces differing only for the order of
% independent steps, referred to as shift equivalence~\cite{CMREHL:AAGT}
% (similar to interchange law~\cite{Mes92} in term rewriting or
% permutation equivalence~\cite{JJL80} in the $\lambda$-calculus).

% In some situations rewriting with linear rules are not sufficiently
% expressive for capturing the dynamic of a system. This is the case
% when a computational step is required not only to consume, preserve
% and generated, but also ``merge'' parts of the state. This happens,
% e.g., in nominal calculi where, as a result of name passing, the
% received name is identified with a local one at the
% receiver~\cite{CVY:ESSPE,Gad07} or in the modelling of bonding in
% biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
% at the core of e-graphs (equality graphs), recently used for
% rewrite-driven compiler optimisations~\cite{WNW:egg}. Rather than
% explicitly merging nodes corresponding to equal expression, the idea
% here - conceptually and, as we will see, also mathematically similar -
% is to maintain an equivalence\todo{P: more precisely a congruence}
% over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
%   vedere se c'\`e qualcosa d'altro}\\
% =========

Technically, to acquire the expressiveness needed for expressing
fusions, i.e., the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories (which capture most
``applications'') have not been systematically explored. ADD REFS

As an initial step in performing this exploration, we observe that for left-linear
rules a number of fundamental properties that play a role in the
corresponding theory of rewriting need to be deeply revised, or plainly do not hold.

Firstly, Church-Rosser theorem does not work out of the box. In
particular, sequential independence between rewriting steps~\cite{}
expressed in terms of the existence of a so-called independence pair
is a canonical and well-understood notion of independence in the
linear case. \todo{P: Too technical? Forse qui e' piu' esplicativo un
  esempio} Whenever two steps are sequential independent
\begin{itemize}
\item they can be switched and 
\item the result of the switch is uniquely determined.
\end{itemize}

Here we observe that, in general, both property can fail: sequential
independent steps could be not switchable and different independent
pairs could exist leading to different switches of the same rewriting
steps.  A refined notion of independence needs to be considered to ensure
that independent rewriting steps can be switched, which is based in the concept
of fillers. The problem was identified
in~\cite{baldan2011adhesivity}, where a subclass of adhesive
categories was identified where independence as induced by a filler
coincides with sequential independence (as induced by an independence
pair). Here it is extended to $\mathcal{M}$-adhesive categories.

We face the problem in the setting of $\mathcal{M}$-adhesive
categories (which capture adhesive and quasi-adhesive categories).
A difficult task, since the most natural idea, put forward in~\cite{baldan2017domains}, 
of limiting sequential independence to those cases in which the switch is possible
and uniquely determined does fall short.
%
In this work we focus on a class of rewriting systems, which we call
\emph{very tame rewriting systems}, in $\mathcal{M}$-adhesive
categories where sequential independence ensures existence and
uniqueness of the switch and argue that this is the right setting for
dealing with left-linear rewriting systems.

On the one hand, we argue that very tame rewriting systems capture most
categories of interest for rewriting. In particular, we single out a
class of left-linear rewriting systems on graph-like structures which
are very tame and we show that rewriting systems on (term)graphs with
equivalences capturing e-graphs are also very tame. As a sanity
check, also linear rewriting systems on $\mathcal{M}$-adhesive
categories are very tame.

On the other hand, we show that an appropriate theory of rewriting can
be developed for very tame rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Local
Curch-Rosser is fully recovered for very tame rewriting systems.

For linear rewriting systems sequential independence of derivations
can be also characterised in terms of the existence of a suitable
isomorphism between the colimit of the derivations, called
consistent permutations (which are also strictly related to the notion
of process). This result, which is folklore for categories of graphs,
is explicitly proved here for $\mathcal{M}$-adhesive categories. Based on such result one can deduce two fundamental consequences:
\begin{enumerate}
\item when switching a derivation consisting of several steps, the result only depends on the overall permutation, not on the order of the switches;
\item independence of rewriting steps is a ``global property'', independent of the position in which these steps are in the derivation.
\end{enumerate}

We show that for left-linear rewriting systems the characterisation of
shift equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2). However property (1) still holds. Instead only a weak form of (2) can be proved. This is intimately related to the possibility of expressing disnjuctive forms of causality.

\todo[inline]{
This fails for left-linear rules where, as observed
in~\cite{baldan2017domains}, for graph rewriting, the possibility of
merging parts of the state leads to a sort of disjunctive causality,
i.e., a rewriting step, say $r_c$ can be enabled by two different
rewriting steps $r_a$ and $r_b$, in a way that in a derivation
$r_a r_b r_c$ steps $r_b$ and $r_c$ are independent since the presence
of $r_a$ in the past is sufficient for enabling $r_a$, while in
$r_b r_c r_a$ , the steps $r_b$ and $r_c$ are not independent. }

@@

  The category $\cat{Graph}$ is tame by the result in ...  Thus,
  sequential independence ensure the existence of a switch $\der{E}$,
  depicted in Fig.~\ref{fi:derE}.
  \begin{figure}
    \input{Figure/derE.tex}
    \caption{The derivation $\der{E}$.}
    \label{fi:derE}
  \end{figure}
\end{example}
\fi

@@

\begin{definition}[Good independence pair]
  \label{def:filler}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also $(i_0, i_1)$ be an independence pair between two direct
  derivations $\dder{D}_0$, $\dder{D}_1$
  as in the solid part of the diagram below
  \[
    \xymatrix@C=18pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} \ar@{.>}@/^.20cm/[ddrr]|(.32)\hole|(.57)\hole_(.7){u_0} & R_0
      \ar@/^.35cm/[drrr]|(.3)\hole^(.2){i_0} \ar[dr]|(.3)\hole_{h_0}
      && L_1 \ar@/_.35cm/[dlll]_(.2){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
      \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} \ar@{.>}@/_.20cm/[ddll]|(.32)\hole|(.57)\hole^(.7){u_1} && R_1 \ar[d]^{h_1} \\
      %
      G_0 &&
      \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_(.3){g_0}&& G_1 &&
      \ar@{>->}[ll]^(.3){f_1} D_1 \ar[rr]_{g_1}&& G_2\\
      %
      & & & & P \ar@/^.3cm/@{>.>}[ull]^{p_0} \ar@/_.3cm/@{.>}[urr]_{p_1}
    }
  \]

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{0,1\}$) and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ into the pullback (see \Cref{prop:tec} for details). 
  % given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{good independence pair} if
  the first two squares depicted  
  below are pushouts and if the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists.
  \[
    \xymatrix{
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
       &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
       & P \ar@{.>}[r]_{q_1} & Q_1
    }
  \]
\end{definition}

% \begin{definition}[Good independence pair]\label{def:filler}
%   Let $(i_0, i_1)$ be an independence pair between two direct
%   derivations $\dder{D}_0$, $\dder{D}_1$ in $(\X, \R)$ be a
%   left-linear DPO rewriting system with $\X$ $\mathcal{M}$-adhesive.
%   \[
%     \xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0
%       \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
%       \ar@/^.35cm/[drrr]|(.3)\hole_(.4){i_0} \ar[dr]|(.3)\hole_{h_0}
%       && L_1 \ar@/_.35cm/[dlll]^(.4){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
%       \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}
%       && P \ar@{>->}[rr]^{p_0} \ar[d]_{p_1}&& D_0\ar[d]^{g_0}\\G_0 &&
%       \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_{g_0}&& G_1 &&
%       \ar@{>->}[ll]^{f_1} D_1 \ar[rr]_{g_1}&& G_2 && D_1
%       \ar@{>->}[rr]_{f_1} && G_1}
%   \]

%   Let also the right square above be a pullback. Let
%   $u_0\colon K_0\to P$ and $u_1\colon K_1\to P$ be the arrows of
%   \Cref{prop:tec}, we will say that $(i_0, i_1)$ is a \emph{good
%     independence pair} if the first and second square below are
%   pushouts and if the third square exists and it is a pushout too.
%   \[\xymatrix{K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
%       \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}&K_1
%       \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0} \\ P
%       \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0&P \ar@{.>}[r]_{q_1} &
%       Q_1}\]
% \end{definition}

We can now prove a Local Church-Rosser Theorem for good independence pairs.

\begin{restatable}[Local Church-Rosser Theorem]{theorem}{thmChurch}
  \label{thm:church}
  Let $(i_0, i_1)$ be a good independence pair
  between $\dder{D}_0$ and $\dder{D}_1$. Then $\dder{D}_0$ and
  $\dder{D}_1$ are switchable.
\end{restatable}
[Proof in \Cref{thmChurch-proof}]

\todo[inline]{Forse non introdurrei neppure la classe dei tame, tanto usiamo solo i very tame, o no?}

@@

The correspondence between sequential independence and switchability
is fundamental.  A very large class of DPO rewriting systems -- along the
lines of \cite{baldan2011adhesivity} -- can be identified where all
independence pairs are good for left-linear rewriting systems. The details are in 
\Cref{app:fill}. Here it is important only to observe that such class includes $\cat{Set}$ and it is closed by comma and functor category constructions. 
As such, it includes essentially all categories (e.g., presheaves over set) that are typically considered for modelling purposes.

%\rem{Non introdurrei una classe, come detto sopra}{
\begin{definition}[Tame rewriting systems]A left-linear DPO rewriting system is \emph{tame} if every independence pair between two direct derivations is good.
\end{definition}

\begin{example}
  \label{ex:diff2}
  In light of \Cref{thm:church}, \Cref{ex:diff1} provides an example
  of an independence pair which is not good, and thus of a non tame
  adhesive rewriting system. Instead the category $\cat{Graph}$ is
  tame by the result in ....
\end{example}

\begin{remark}
  By \Cref{thm:church} the existence of a good independence pair
  entails switchability which, in turn, entails weak sequential
  independence by construction. Tame rewriting systems are exactly
  those rewriting systems in which these three notions coincide.
\end{remark}
%}

@@

A source of tame left-linear DPO rewriting systems is given by the linear ones, as shown by the following proposition.

\begin{restatable}{proposition}{propEqui}
  \label{prop:equi}
  Every linear DPO rewriting system $(\X, \R)$ is tame.
\end{restatable}
[Proof in \Cref{propEqui-proof}]

\subsection{Very tame rewriting systems}\label{subsec:verytame}

@@

\begin{definition}[Very tame rewriting systems]
A left-linear DPO rewriting system $(\X, \R)$ is \emph{very tame} if it is tame and, for every derivation $\der{D}:=\{\dder{D}_{i}\}_{i=0}^1$, there is at most one independence paire between $\dder{D}_0$ and $\dder{D}_1$.
\end{definition}

>>>>

% fake author for anonymous submission
\author{Alan Turing} 
{Department of Mathematics, Somewhere}
{alan@turing.com}{}{}

% \author{Paolo Baldan} 
% {Department of Mathematics, University of Padova, Italy}
% {baldan@math.unipd.it}{}{}
% \author{Davide Castelnovo}
% {Department of Mathematics, University of Padova, Italy}
% {davide.castelnovo@math.unipd.it}{}{}
% \author{Andrea Corradini}
% {Department of Computer Science, University of Pisa, Italy}
% {andrea.corradini@unipi.it}{}{}

% \author{Fabio Gadducci}
% {Department of Computer Science, University of Pisa, Italy}
% {fabio.gadducci@unipi.it}{}{}

@@

\begin{abstract}
  When two sequences of steps executed by a computational device can
  be considered equivalent?  This is a relevant question if one
  considers concurrent and distributed systems, since the independence
  of two steps means that they could be executed in any order, and
  potentially in parallel.
  % 
  We investigate the issue in the context of adhesive categories, a
  very general instance of the DPO approach to rewriting. More
  precisely, we focus on left-linear rules, which not only consume,
  preserve and generate entities, as linear rules do, but may
  ``merge'' parts of the state.
  % 
  An apparently minimal, yet technically hard enhancement, 
  % 
  % We refine the established notion of sequential independence
  % between steps, and ...
  since the canonical notion of independence used for linear rewriting
  systems does not work out of the box, but it requires the
  restriction to well-behaved classes of left-linear rules.
  %
  Moreover, a standard characterisation of independence that -- in the
  linear case -- allows one to derive a number of properties,
  essential in the development of a theory of concurrency, no longer
  holds, and this requires a substantial revision of the
  theory.
  %
  The paper performs an in-depth study of the notion of independence
  for left-linear rules: it introduces a novel characterisation of
  independence, identifies well-behaved classes of left-linear
  rewriting systems and provides some fundamental results including a
  Church-Rosser property and the existence of canonical equivalence
  proofs for concurrent computations. These results properly extends
  the class of formalisms that can be modelled in the adhesive
  framework.
\end{abstract}

@@

\item \emph{Strong Enforcing rewriting systems}\\ An
  independence pairs induced by a filler is called \emph{strong} A
  rewriting system where all indepedence pairs are strong is called
  \emph{strong enforcing}

  NOTE: the categories in which
  condition $B$ from~\cite{baldan2011adhesivity} identifies a large
  class of (quasi)adhesive categories where left-linear systems are all
  strong enforcing. Adapted here to M-adhesivity.

\item \emph{Well Switching rewriting
    systems}\\ Strong Enforcing and existing indepence pairs are unique (hence
  switching is unique) (see later for examples)
\end{itemize}

@@

This part is dealt with for well switching rewriting systems and depends on this assumption. Note that
\begin{itemize}
\item Linear rewriting systems on adhesive categories are well switching
\item Left-linear rewritint systems on graph-like categories with generalisation of (i) no isolated nodes in $L$ and (ii) right leg injective on nodes is well switching
\item Graphs with equivalences (Egg) with left-leg regular mono, right leg mono is well switching
\end{itemize}

@@

So far, most of the theoretical results focussed on what are called \emph{linear rules}: 
intuitively, a step is required only to consume, preserve and generate entities. 
However, in some situations it is also necessary to ``merge'' parts of the state. 
This happens already in the context of string diagram mentioned above, but also 
in graphical implementations of nominal calculi where, as a result of name passing, 
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the visual modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimisations~\cite{WNW:egg}: rather than
explicitly merging nodes corresponding to equal expression, the idea
here -- conceptually and, as we will see, also mathematically similar --
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

Technically, to acquire the expressiveness needed for expressing
fusions, i.e., the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when devising a theory of indepedence and
concurrency for rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories
%(which capture most ``applications'') 
have not been systematically
explored~\cite{Ehrig1976,EHP:BRfTToHLRS,baldan2011adhesivity} \todo{P: see also \cite{HMP:DPORADD,BHK:CTNL} ... verificare e aggiungere}

Here we perform this exploration working at the level of $\mathcal{M}$-adhesive catgories, which encompass ...

As an initial step in performing this exploration, we observe that for
left-linear rules a number of fundamental properties that play a role
in the corresponding theory of rewriting need to be deeply revised, or
plainly do not hold.
%
Firstly, the notion of sequential independence and Church Rosser
theorem, ensuring that sequential independent steps can be performed
in reverse order, does not hold out of the box. Even worse, given two
independent steps it might be possible to switch them in several
different ways. As we will observe, this might prevent the development
of a sensible theory of concurrency as switches are used as a basis
witch equivalence of derivation sequences, identifying rewriting
sequences of steps which differs only for the order of independent
steps.

We single a well-behaved class of left-linear rewriting
systems referred to as \emph{well-switching rewriting systems}, where sequential independence ensures existence and uniqueness of the
switch, and argue that this is the right setting for dealing with
left-linear rewriting systems.

On the one hand, we argue that they capture most categories of
interest for rewriting. In particular, we single out a class of
left-linear rewriting systems on graph-like structures which are very
well-switchable and we show that rewriting systems on graphs with
equivalences, similar in spirit to e-graphs are also
well-switchable. As a sanity check, also linear rewriting systems on
$\mathcal{M}$-adhesive categories are well switching.

On the other hand, we show that an appropriate theory of rewriting can
be developed for well switching rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Church-Rosser
is fully recovered for well-switchable rewriting systems.

There is, however, a further, serious problem.  For linear
rewriting systems s, can be characterised in terms of the
existence of what is called a consistent permutation~\cite{xxx}
(roughly, an isomorphism between their colimits). This result, which
is folklore for various categories of graphs, is explicitly proved
here for $\mathcal{M}$-adhesive categories.
%
The characterisation of switch equivalence based on consistent
permutations is the key to derive some
properties for linear rewriting systems, namely
\begin{enumerate}
\item \emph{confluence of switching}: when transforming derivation
  sequences into equivalent ones by sqwitching independent steps, the
  result not depends on the order in which switchings are applied, but
  only on the resulting permutation;

\item \emph{globality of independence}: if two independent steps are
  moved forward or backward due to the application of switchings to a
  derivation, they remain independent.
\end{enumerate}
Properties (1) and (2) allow one to deduce a canonical form for equivalence proofs between derivations seen as concurrent computations.

Unfortunately, for left-linear rewriting systems the characterisation
of switch equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2).
%
This leads to the question asking whether the properties above hold.
We show that actually much of the theory can be retained: confluence
of switching still holds, while globality of independence holds in a
weak form, conceptually linked to the fact that fusions introduce a
form of disjunctive causality. Finally, a canonical form for switching
sequences can also be recovered.

% \todo[inline]{
% This fails for left-linear rules where, as observed
% in~\cite{baldan2017domains}, for graph rewriting, the possibility of
% merging parts of the state leads to a sort of disjunctive causality,
% i.e., a rewriting step, say $r_c$ can be enabled by two different
% rewriting steps $r_a$ and $r_b$, in a way that in a derivation
% $r_a r_b r_c$ steps $r_b$ and $r_c$ are independent since the presence
% of $r_a$ in the past is sufficient for enabling $r_a$, while in
% $r_b r_c r_a$ , the steps $r_b$ and $r_c$ are not independent.
% }

% \todo[inline]{
% DA RIUSARE\\
% Firstly, Church-Rosser theorem does not work out of the box. In
% particular, sequential independence between rewriting steps~\cite{}
% expressed in terms of the existence of a so-called independence pair
% is a canonical and well-understood notion of independence in the
% linear case. \todo{P: Too technical? Forse qui e' piu' esplicativo un
%   esempio} Whenever two steps are sequential independent
% \begin{itemize}
% \item they can be switched and 
% \item the result of the switch is uniquely determined.
% \end{itemize}
% Here we observe that, in general, both property can fail: sequential
% independent steps could be not switchable and different independent
% pairs could exist leading to different switches of the same rewriting
% steps.  A refined notion of independence needs to be considered to ensure
% that independent rewriting steps can be switched, which is based in the concept
% of fillers. The problem was identified
% in~\cite{baldan2011adhesivity}, where a subclass of adhesive
% categories was identified where independence as induced by a filler
% coincides with sequential independence (as induced by an independence
% pair). Here it is extended to $\mathcal{M}$-adhesive categories.
% }

%   First of all, we will provide a refined notion of independence,
%   which is based in the concept of fillers.

% Then, we move to restrict the class of admissible left-linear rewriting systems,
% always working at the level of $\mathcal{M}$-adhesive categories.
% Unfortunately, the most natural idea, put forward in~\cite{baldan2017domains}, 
% of limiting sequential independence to those cases in which the switch is possible
% and uniquely determined does fall short.

@@

  The category $\cat{Graph}$ is strong enforcing by the result in ...  Thus,
  sequential independence ensure the existence of a switch $\der{E}$,
  depicted in Fig.~\ref{fi:derE}.
  \begin{figure}
    \input{Figure/derE.tex}
    \caption{The derivation $\der{E}$.}
    \label{fi:derE}
  \end{figure}
\end{example}
\fi

@@

\begin{definition}[strong independence pair]
  \label{def:filler}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also $(i_0, i_1)$ be an independence pair between two direct
  derivations $\dder{D}_0$, $\dder{D}_1$
  as in the solid part of the diagram below
  \[
    \xymatrix@C=18pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} \ar@{.>}@/^.20cm/[ddrr]|(.32)\hole|(.57)\hole_(.7){u_0} & R_0
      \ar@/^.35cm/[drrr]|(.3)\hole^(.2){i_0} \ar[dr]|(.3)\hole_{h_0}
      && L_1 \ar@/_.35cm/[dlll]_(.2){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
      \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} \ar@{.>}@/_.20cm/[ddll]|(.32)\hole|(.57)\hole^(.7){u_1} && R_1 \ar[d]^{h_1} \\
      %
      G_0 &&
      \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_(.3){g_0}&& G_1 &&
      \ar@{>->}[ll]^(.3){f_1} D_1 \ar[rr]_{g_1}&& G_2\\
      %
      & & & & P \ar@/^.3cm/@{>.>}[ull]^{p_0} \ar@/_.3cm/@{.>}[urr]_{p_1}
    }
  \]

  Consider the pullback of $g_0$ and $f_1$, getting $p_i : P \to D_i$
  ($i \in \{0,1\}$) and the mediating arrows $u_0\colon K_0\to P$ and
  $u_1\colon K_1\to P$ into the pullback (see \Cref{prop:tec} for details). 
  % given by \Cref{prop:tec} .
  %
  We will say that $(i_0, i_1)$ is a \emph{strong independence pair} if
  the first two squares depicted  
  below are pushouts and if the pushout of $r_1 : K_1 \to R_1$ and $u_1 : K_1 \to P$ exists.
  \[
    \xymatrix{
      K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
      \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}
       &K_1 \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0}
      \\
      %
      P \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0
       & P \ar@{.>}[r]_{q_1} & Q_1
    }
  \]
\end{definition}

% \begin{definition}[strong independence pair]\label{def:filler}
%   Let $(i_0, i_1)$ be an independence pair between two direct
%   derivations $\dder{D}_0$, $\dder{D}_1$ in $(\X, \R)$ be a
%   left-linear DPO rewriting system with $\X$ $\mathcal{M}$-adhesive.
%   \[
%     \xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0
%       \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
%       \ar@/^.35cm/[drrr]|(.3)\hole_(.4){i_0} \ar[dr]|(.3)\hole_{h_0}
%       && L_1 \ar@/_.35cm/[dlll]^(.4){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
%       \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}
%       && P \ar@{>->}[rr]^{p_0} \ar[d]_{p_1}&& D_0\ar[d]^{g_0}\\G_0 &&
%       \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_{g_0}&& G_1 &&
%       \ar@{>->}[ll]^{f_1} D_1 \ar[rr]_{g_1}&& G_2 && D_1
%       \ar@{>->}[rr]_{f_1} && G_1}
%   \]

%   Let also the right square above be a pullback. Let
%   $u_0\colon K_0\to P$ and $u_1\colon K_1\to P$ be the arrows of
%   \Cref{prop:tec}, we will say that $(i_0, i_1)$ is a \emph{strong
%     independence pair} if the first and second square below are
%   pushouts and if the third square exists and it is a pushout too.
%   \[\xymatrix{K_0 \ar[r]^{r_0} \ar[d]_{u_0}& R_0 \ar[d]^{i_0} & K_1
%       \ar@{>->}[r]^{l_1} \ar[d]_{u_1}& L_1 \ar[d]^{i_1}&K_1
%       \ar[r]^{r_1} \ar[d]_{u_1}& R_1\ar@{.>}[d]^{j_0} \\ P
%       \ar[r]_{p_1} & D_1 &P \ar[r]_{p_0} & D_0&P \ar@{.>}[r]_{q_1} &
%       Q_1}\]
% \end{definition}

We can now prove a Local Church-Rosser Theorem for strong independence pairs.

\begin{restatable}[Local Church-Rosser Theorem]{theorem}{thmChurch}
  \label{thm:church}
  Let $(i_0, i_1)$ be a strong independence pair
  between $\dder{D}_0$ and $\dder{D}_1$. Then $\dder{D}_0$ and
  $\dder{D}_1$ are switchable.
\end{restatable}
[Proof in \Cref{thmChurch-proof}]

\todo[inline]{Forse non introdurrei neppure la classe dei strong enforcing, tanto usiamo solo i well switching, o no?}

@@

The correspondence between sequential independence and switchability
is fundamental.  A very large class of DPO rewriting systems -- along the
lines of \cite{baldan2011adhesivity} -- can be identified where all
independence pairs are strong for left-linear rewriting systems. The details are in 
\Cref{app:fill}. Here it is important only to observe that such class includes $\cat{Set}$ and it is closed by comma and functor category constructions. 
As such, it includes essentially all categories (e.g., presheaves over set) that are typically considered for modelling purposes.

%\rem{Non introdurrei una classe, come detto sopra}{
\begin{definition}[Strong Enforcing rewriting systems]A left-linear DPO rewriting system is \emph{strong enforcing} if every independence pair between two direct derivations is strong.
\end{definition}

\begin{example}
  \label{ex:diff2}
  In light of \Cref{thm:church}, \Cref{ex:diff1} provides an example
  of an independence pair which is not strong, and thus of a non strong enforcing
  adhesive rewriting system. Instead the category $\cat{Graph}$ is
  strong enforcing by the result in ....
\end{example}

\begin{remark}
  By \Cref{thm:church} the existence of a strong independence pair
  entails switchability which, in turn, entails weak sequential
  independence by construction. Strong Enforcing rewriting systems are exactly
  those rewriting systems in which these three notions coincide.
\end{remark}
%}

@@

A source of strong enforcing left-linear DPO rewriting systems is given by the linear ones, as shown by the following proposition.

\begin{restatable}{proposition}{propEqui}
  \label{prop:equi}
  Every linear DPO rewriting system $(\X, \R)$ is strong enforcing.
\end{restatable}
[Proof in \Cref{propEqui-proof}]

\subsection{well switching rewriting systems}\label{subsec:verytame}

@@

\begin{definition}[well switching rewriting systems]
A left-linear DPO rewriting system $(\X, \R)$ is \emph{well switching} if it is strong enforcing and, for every derivation $\der{D}:=\{\dder{D}_{i}\}_{i=0}^1$, there is at most one independence paire between $\dder{D}_0$ and $\dder{D}_1$.
\end{definition}
